---
title       : Ellipse tangent
subtitle    : Coursera - Data Products class
author      : Guillaume Bourgault
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## Project motivation
I recently started to learn how to draw, and one of the things I find hard is to draw realistic ellipses.  They come up often, because there are so many round objects in our surroundings: glasses, candles, chair bars, lamps, etc.  When they are seen at an angle, in 2D, they appear as ellipses.  

To help me, I figured out this procedure:

1. Find what is the width-to-height ratio of the ellipse
2. Draw the 4 extremities of the ellipse
3. Find where is the point, relative to the extremities, where the tangent is at 45 degrees.
4. Draw a very short segment at a 45 degree angle relative to the vertical.  
5. Play connect the dots with a smooth curve.  

---
## Drawing an ellipse: not that easy!
After doing the 5 steps described, things look like this on my sketching pad.  
I was always surprized to see how close to the tip the 45 degree tangent point was.  
I decided to investigate the relationship between the position of this point as a function of the width-to-ratio and the angle of the tangent chosen.  
```{r, echo = FALSE, warning=FALSE, quietly=TRUE}
library(aspace)
library(dplyr)
ellipse_y <- function(a, x) {
  return(sqrt(1-(x/a)**2))
}
x_y_from_angle <- function(a, theta) {
  x <- tan(as_radians(theta))*a*((1/a**2)+(tan(as_radians(theta)))**2)**-.5
  y <- ellipse_y(a, x)
  return(list(x = x, y = y))
}
theta = 45
a = 2.
x_y <- x_y_from_angle(a, theta)
df1 <- data.frame(x = c(a, x_y$x, 0, -x_y$x, -a))
df1 <- mutate(df1, y = ellipse_y(a, x))
df2 <- data.frame(x = c(a, x_y$x, 0, -x_y$x, -a))
df2 <- mutate(df2, y = -ellipse_y(a, x))
df <- rbind(df1, df2)
plot(df, asp = 1, axes=FALSE, frame.plot=TRUE, xlab = '', ylab = '')
```

---
## A bit of math
The equation of an ellipse, centered at (0,0) is:
$$\frac{x^2}{a^2} + \frac{y^2}{b^2} = 1$$
To simplify the equation, we assume b = 1.  The first step is to express y = f(x).  
$$y = f(x) = \sqrt{1-\frac{x^2}{a^2}}$$
We then find the derivative of y = f(x), and equal it to tan(theta).
$$\frac{\mathrm{d} y}{\mathrm{d} x} = -\frac{x}{a^{2}\sqrt{1-\frac{x^{2}}{a^{2}}}} = tan(\theta)$$ 
It is now possible to express x as a function of theta and a:
$$x = \frac{a\cdot tan(\theta)}{\sqrt{\frac{1}{a^{2}}+tan^{2}\theta}}$$

---
## Shiny app
The application contains two sliders, one for a and one for theta.  
The output uses a reactive plotting feature, yealding a graph that updates each time a or theta are changed on the sliders.  Have fun!

```{r, echo = FALSE, warning=FALSE, quietly=TRUE}
ellipse_with_tangent <- function(a, theta) {
  #plots the ellipse and a tangent.  
  ellipse_points <- ellipse(a)
  intercept <- intercepts(a, theta)
  max_y = intercept[1, 'y']*1.05
  plot(ellipse_points, type = 'l', asp = 1, 
       xlab = '', 
       ylab = '', 
       xlim = c(-max_a, max_a), 
       ylim = c(-max_y, max_y), 
       axes = TRUE)
  lines(intercept)
  lines(intercept[c(1,2,2),'x'], intercept[c(2,2,3),'y'])
  lines(c(-max_a, max_a), c(0,0))
  lines(c(0,0), c(-max_y, max_y))
  title('An ellipse and its tangent')
}
intercepts <- function(a, theta) {
  #Calculates where the tangent of an ellipse intercepts the x and y axis. 
  x_y <- x_y_from_angle(a, theta)
  x <- c(0, x_y$x, x_y$x + x_y$y/tan(as_radians(theta)))
  y <- c(x_y$y + x_y$x*tan(as_radians(theta)), x_y$y, 0)
  return(data.frame(x, y))
}
ellipse <- function(a) {
  #builds the points on an ellipse of width-to-ratio a.  
  nb_points <- 1000
  df1 <- data.frame(list(x = seq(-a, a, a/nb_points)))
  df1 <- mutate(df1, y = ellipse_y(a, x))
  df2 <- data.frame(list(x = seq(-a, a, a/nb_points)))
  df2 <- mutate(df2, y = -ellipse_y(a, x))
  df <- rbind(df1, df2)
  return(df)
}
ellipse_with_tangent <- function(a, theta) {
  #plots the ellipse and a tangent.  
  ellipse_points <- ellipse(a)
  intercept <- intercepts(a, theta)
  max_y = intercept[1, 'y']*1.05
  plot(ellipse_points, type = 'l', asp = 1, 
       xlab = '', 
       ylab = '', 
       xlim = c(-max_a, max_a), 
       ylim = c(-max_y, max_y), 
       axes = TRUE)
  lines(intercept)
  lines(intercept[c(1,2,2),'x'], intercept[c(2,2,3),'y'])
  lines(c(-max_a, max_a), c(0,0))
  lines(c(0,0), c(-max_y, max_y))
  title('An ellipse and its tangent')
}
max_a = 4.
ellipse_with_tangent(a, theta)
```